Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-19 04-analisis.qmd
analysis & discussion

@fig-correlation presents a correlation matrix of the main variables analyzed, using data from the first wave. In this matrix, the correlations vary between low and moderate values. The correlations between justification of educational inequality with the two covid variables are positive, but low. In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.34, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.29, p<.01; _r_=0.24, p<.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.18, p<.01; _r_=-0.08, p<.01). Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.76, p<.01).

@@

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05). Nevertheless, in the last waves of 2022 and 2023 there is a clear increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time. On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.

Models 3 and 4 adds the two variables about the Covid context, where we can see that those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01). On the other hand, those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.038, p<.05).

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.055, p<.001). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent. Inequality perception is added in Model 7, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.

Model 8 adds the socioeconomic variables to test hypothesis 7. Educational level and household per capita income were added, but none of them had a significant effect. However, in the next models, controlling for all variables, household per capita income had a positive significant effect ($\beta$=0.001, p<.05). In addition, in this model the demographic control variables were added, where being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001), and being women (compared to men) had a negative and significant effect ($\beta$=-0.08, p<.001). Subjective social status and age had no significant effect when controlling for all the variables analysed. The full models can be seen in the appendix (see @tbl-multilevel-full).

@@

@@

In this last part of the analysis, we test hypotheses about the effect of COVID-19 on changes in the relationship between meritocracy and justification of educational inequality. In hypothesis 5, we proposed that the association between the perception of meritocracy and justification of inequality is mitigated in times of crisis, as meritocratic ideals might have been weakened due to critical situations associated with the COVID-19 pandemic. We test this hypothesis through interaction effects, shown in @tbl-interact. The first model is shown as the base model. It is the same as model 8 in @tbl-multilevel, but only shows the variables involved in the interaction for space reasons (all other variables are controlled). Model 9 adds the interaction between the perception of talent-based meritocracy and whether state benefits were received during the pandemic. Meanwhile, model 10 does the same, but now, it is for the perception of effort-related meritocracy. As can be seen, both variables show a negative significant interaction with receiving state benefits ($\beta_{talent}$=-0.07, p<.01; $\beta_{effort}$=-0.05, p<.05), which means that, on average, the association between the perception of talent-based meritocracy and the justification of inequality is lower by 0.07 points for those who received state benefits and the association between the perception of effort-based meritocracy and the justification of inequality is lower by 0.05 points for those who received state benefits.

@@

To further explore this interaction, the interaction between receiving state benefits and the perception of meritocracy based on talent can be visualized in @fig-interact-talent. This figure shows that the positive effect of perceiving greater meritocracy based on talent on the justification of educational inequality is smaller for those who received state benefits during the pandemic.

>>>>

@fig-correlation presents a correlation matrix of the main variables analyzed, using data from the first wave. In this matrix, the correlations vary between low and moderate values. The correlations between justification of educational inequality with the two COVID variables are positive, but low. In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.34, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.29, p<.01; _r_=0.24, p<.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.18, p<.01; _r_=-0.08, p<.01). Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.76, p<.01).

@@

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05). Nevertheless, in the last waves of 2022 and 2023 there is a clear increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time. On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement points.

Models 3 and 4 adds the two variables about the pandemic context, where we can see that those who received state benefits during this period justify less inequality in education compared to those who did not receive them. Although the effect size is small, it remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01). On the other hand, those who believe that in this context the economy is more important than the health of the population justify more inequality in education($\beta$=0.038, p<.05, Model 8).

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.055, p<.001). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent. Regarding inequality perception (added in Model 7), it depicts a negative association with the justification of educational inequality, remaining stable when controlling for the rest of the variables.

Model 8 adds the socioeconomic variables to test hypothesis 7. Although educational level and household per capita income are not significant in this model, controlling for all variables in the next models, household per capita income depicts a positive effect ($\beta$=0.001, p<.05). The inclusion of other controls as being right-wing (compared to being left-wing) shows a positive association with justification of educational inequality ($\beta$=0.27, p<.001), whereas being a women (compared to men) had a negative  effect ($\beta$=-0.08, p<.001). Subjective social status and age had no significant effect when controlling for all the variables analyzed. The full model specification can be seen in the appendix (see @tbl-multilevel-full).

@@

In this last part of the analysis we test hypotheses about the effect of COVID-19 on changes in the relationship between meritocracy and justification of educational inequality. In hypothesis 5, we proposed that the association between the perception of meritocracy and justification of inequality is mitigated in times of crisis, as meritocratic ideals might have been weakened due to critical situations associated with the COVID-19 pandemic. We test this hypothesis through interaction effects, shown in @tbl-interact. The first model is shown as the baseline model. It is the same as model 8 in @tbl-multilevel, but only shows the variables involved in the interaction for the sake of space (all other variables are included). Model 9 adds the interaction between the perception of talent-based meritocracy and whether state benefits were received during the pandemic, whereas Model 10 considers the perception of effort-related meritocracy for the interaction. Both variables show a negative significant interaction with receiving state benefits ($\beta_{talent}$=-0.07, p<.01; $\beta_{effort}$=-0.05, p<.05), meaning that for those who received state benefits the association between the the justification of inequality and the perception of effort-based meritocracy lower by 0.05 points, and by 0.07 points regarding the perception of talent-based meritocracy. This last interaction is depicted in @fig-interact-talent, showing that the positive association between the perception of talent-based meritocracy and the justification of educational inequality is smaller for those who received state benefits during the pandemic.

@@

@@

### 2025-01-19 05-discussion.qmd
analysis & discussion

Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic. In general, our results showed mixed evidence about our hypotheses. Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal. We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens. In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic. As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises. As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies. A closer look to the political environment during the last year in the country could give additional hints. Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly. This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chiles_2023;@sazo_chile_2023].

Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$). In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification. Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$). Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.

The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$). At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent. Our results are favorable for the effort dimension but not confirmed for the talent dimension. In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent. This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.

Our first hypothesis was supported, as we anticipated that the relationship between perceptions of meritocracy and the justification of educational inequality would tend to diminish over time. We based this on questions that related directly to experiences during the pandemic, such as receiving economic benefits during the lockdown period. Our findings revealed that those who received benefits justified educational inequality less than those who did not. Moreover, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received benefits. This evidence partially supports our temporal change hypothesis ($H5$). In other words, the influence of meritocracy on justifying inequality appears to lessen, which aligns with our expectations given the social context in the post-COVID era. One possible interpretation is that, during times of increased risk, people may become more skeptical of narratives emphasizing individual effort over collective efforts. This shift in perspective might also foster stronger solidarity to protect individual and societal wellbeing. However, further studies are necessary to explore this phenomenon using more specific conceptual and methodological approaches.

Among the limitations of our study, we can mention at least three. First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies. In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions. Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital. In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard. Finally, there are limitations involving the longitudinal dimension of the data used. Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey. Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.

The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates. In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies. For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.  Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions. Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.

>>>>

Our research sought to examine the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic. In general, our results showed mixed evidence about our hypotheses. We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens. In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the post-pandemic measurement points. Besides, our findings revealed that those who received benefits justified educational inequality less than those who did not. Still, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises. As during this very same period in Chile there were two (failed) constitutional processes, disentangling different forces driving public opinion becomes cumbersome and deserves further qualitative and quantitative studies. Along this line, some recent studies have argued that the election of a far left-wing constitutional assembly in the first process generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly. This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chiles_2023;@sazo_chile_2023].

The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education. At first look, the results are consistent with the hypothesis; however, some nuances are worth attending as the effort dimension prevails over the talent dimension. In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent. This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality. Furthermore, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received state benefits, which aligns with our expectations given the social context in the post-COVID era. One possible interpretation is that, during times of increased risk, people may become more skeptical of narratives emphasizing individual effort over collective efforts. This shift in perspective might also foster stronger solidarity to protect individual and societal wellbeing. However, further studies are necessary to explore this phenomenon using more specific conceptual and methodological approaches.

Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality. In this regard, our results show a positive relationship between income and the justification of inequality, which would align with our argument that rational interests of the better-off that would lead to a larger inequality justification. Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude. Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.

Among the limitations of our study, we can mention at least three. First, the operationalization of justification of educational inequality is rather limited. We counted only on a single item for this in this study, and further work is needed in order to assess the validity and reliability of measuring such construct. Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also taking into account the role of structural aspects such as family status and social capital. Due to data limitations such dimensions were not included here, therefore providing a partial perspective on the areas of perceived meritocracy. Finally, there are limitations involving the longitudinal dimension of the data used. Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey. Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.

Regarding future studies, the capabilities of the ELSOC longitudinal database are not limited to micro-level estimates. In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level. For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.  Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions. Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
